---
title: "exdqlm — Extended Dynamic Quantile Linear Models"
output:
  rmarkdown::github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%",
  warning   = FALSE,
  message   = FALSE
)
```

<!-- badges: start -->

[![R-CMD-check](https://github.com/AntonioAPDL/exdqlm/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/AntonioAPDL/exdqlm/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

`exdqlm` fits **dynamic quantile** state-space models using the **extended Asymmetric Laplace** (exAL) error family. It targets time-series applications where the goal is to model **conditional quantiles** across time with a standard state-space specification (design/evolution matrices and a state vector).

**What’s new in v0.5.0.** We add **posterior predictive synthesis** across multiple quantile-specific models via `exdqlm_synthesize_from_draws()`. The method performs **isotonic adjustment** on anchors, **distributional alignment**, **piecewise-linear blending**, and optional **monotone rearrangement** to produce a single coherent predictive distribution per time point. LDVB (`exdqlmLDVB`) and IS-VB remain available; examples below keep the **pure-R** path for portability.

> **Terminology.** We use **exAL** for the extended Asymmetric Laplace family used here. It extends the standard AL by adding a **skewness** parameter; the standard AL is the special case with zero skewness.

## Installation

CRAN (when available):

```r
install.packages("exdqlm")
```

Development (GitHub):

```r
# install.packages("pak")
pak::pak("AntonioAPDL/exdqlm")
```

## Quick start (≤ 10 lines, LDVB)

Local-level model at a **single quantile** (the median). We fix **scale** and **skewness** to keep it fast and stable for CRAN; we keep the **pure-R** path.

```{r quick-start-ldvb}
set.seed(1)
library(exdqlm)

T      <- 120
state  <- cumsum(rnorm(T, sd = 0.2))
y      <- state + rnorm(T, sd = 1.0)

model  <- list(FF = matrix(1), GG = matrix(1), m0 = 0, C0 = 100)
options(exdqlm.use_cpp_kf = FALSE, exdqlm.use_cpp_samplers = FALSE)

fit_ld <- exdqlmLDVB(
  y = y, p0 = 0.5, model = model, df = 0.98, dim.df = 1,
  fix.sigma = TRUE, sig.init = 1.0,
  fix.gamma = TRUE, gam.init = 0.0
)

tail(fit_ld$diagnostics$elbo, 3)
```

## Synthesis workflow (at a glance)

* **Fit** models at a small set of anchor quantiles (e.g., 0.25, 0.5, 0.75).
* **Isotonic adjustment** at the anchors to reduce crossings.
* **Distributional alignment** to co-locate supports.
* **Piecewise-linear blend** in the probability scale to fill gaps.
* **Optional monotone rearrangement** to enforce global monotonicity.

## What’s new in v0.5.0

* **Posterior predictive synthesis**: `exdqlm_synthesize_from_draws()` combines multiple quantile-model draws into one coherent predictive distribution per time point.
* **Diagnostics**: examples outline crossing share, RMSE at anchors, and monotonicity checks.
* **No breaking changes**: LDVB/IS-VB interfaces unchanged; C++ bridges remain opt-in.

### Runtime options (summary)

| Option                    | Default | Effect                           | Use when…                                |
| ------------------------- | :-----: | -------------------------------- | ---------------------------------------- |
| `exdqlm.use_cpp_kf`       |  FALSE  | C++ Kalman filter bridge         | you have compilers/OpenMP and want speed |
| `exdqlm.use_cpp_samplers` |  FALSE  | C++ samplers for posterior draws | same as above; keep OFF on CRAN/examples |

Set with:

```r
options(exdqlm.use_cpp_kf = TRUE)
options(exdqlm.use_cpp_samplers = TRUE)
```

## Minimal examples (CRAN-safe)

### 1) LDVB on built-in data (tiny slice)

Trend + seasonality + one regressor (`nino34`). **Note**: `FF` for the regressor is `1 × T`. Combine components **pairwise**.

```{r builtins-ldvb}
set.seed(2)
T <- 150
y <- log(BTflow[seq_len(T)])
x <- nino34[seq_len(T)]

trend.comp <- polytrendMod(order = 1, m0 = 0, C0 = 1)
seas.comp  <- seasMod(p = 12, h = 1, C0 = diag(1, 2))

# 1-d regressor block (explicit 1 x T design)
reg.comp <- list(m0 = 0, C0 = 1, FF = matrix(x, nrow = 1), GG = matrix(1))

# combine pairwise
base.mod <- combineMods(trend.comp, seas.comp)
model    <- combineMods(base.mod, reg.comp)

# one discount per block: (trend, seasonal[2-d], reg)
df     <- c(1.00, 0.98, 1.00)
dim.df <- c(1,       2,   1)

options(exdqlm.use_cpp_kf = FALSE, exdqlm.use_cpp_samplers = FALSE)

fit_ld <- exdqlmLDVB(
  y = y, p0 = 0.5, model = model,
  df = df, dim.df = dim.df,
  fix.sigma = TRUE, sig.init = 0.2,
  fix.gamma = TRUE, gam.init = 0.0
)

# quick checks
tail(fit_ld$diagnostics$elbo, 2)
dim(fit_ld$theta.out$sm)  # state-dimension x time
```

### 2) Multi-quantile LDVB + synthesis (three anchors)

We fit **three** anchor quantiles, collect their posterior predictive draws, then call `exdqlm_synthesize_from_draws()`.
We also compute simple diagnostics and show a tiny zoom plot.

```{r synthesis}
set.seed(3)
options(exdqlm.use_cpp_kf = FALSE, exdqlm.use_cpp_samplers = FALSE)

p_grid <- c(0.25, 0.5, 0.75)

fit_one <- function(p0) {
  exdqlmLDVB(
    y = y, p0 = p0, model = model,
    df = df, dim.df = dim.df,
    fix.sigma = TRUE, sig.init = 0.2,
    fix.gamma = TRUE, gam.init = 0.0,
    verbose = FALSE
  )
}

fits <- lapply(p_grid, fit_one)
draws_list <- lapply(fits, function(f) f$samp.post.pred)  # list of T×ns (or ns×T)

# --- Synthesize (uses isotonic adj. -> alignment -> blending -> optional rearrangement)
syn <- exdqlm_synthesize_from_draws(
  draws_list = draws_list,
  p          = p_grid
  # defaults perform the full workflow; see ?exdqlm_synthesize_from_draws
)

# Small helper for row-quantiles
rowQ <- function(M, prob) {
  if (is.null(M)) return(NULL)
  if (nrow(M) != T) M <- t(M)
  if (requireNamespace("matrixStats", quietly = TRUE)) {
    as.numeric(matrixStats::rowQuantiles(M, probs = prob, na.rm = TRUE))
  } else {
    apply(M, 1L, quantile, probs = prob, na.rm = TRUE)
  }
}

# --- Diagnostics: pre-synthesis crossing share + RMSE @ anchors (syn vs anchors)
# anchor quantiles from each model
Q_anchor <- lapply(seq_along(p_grid), function(i) rowQ(draws_list[[i]], p_grid[i]))
Q_anchor <- do.call(cbind, Q_anchor)  # T x length(p_grid)

# crossing share BEFORE synthesis (any violation at a time index)
viol <- apply(Q_anchor, 1L, function(v) any(diff(v) < 0))
crossing_share <- mean(viol)  # in [0,1]

# synthesized anchor quantiles (from syn draws)
if (nrow(syn$draws) != T) syn$draws <- t(syn$draws)  # ensure T x ns
Q_syn_at_p <- sapply(p_grid, function(p0) rowQ(syn$draws, p0))
rmse_anchor <- sqrt(colMeans((Q_syn_at_p - Q_anchor)^2))

c(
  crossing_share_pre  = rounding <- round(crossing_share, 3),
  rmse_at_anchors_25  = round(rmse_anchor[1], 4),
  rmse_at_anchors_50  = round(rmse_anchor[2], 4),
  rmse_at_anchors_75  = round(rmse_anchor[3], 4)
)

# --- Minimal zoom plot for the synthesized predictive (last 40 points)
pp_plot_zoom <- function(y, samples, p0 = 0.5, last_n = 40L) {
  if (nrow(samples) != length(y)) samples <- t(samples)
  idx <- (length(y) - last_n + 1L):length(y)
  S   <- samples[idx, , drop = FALSE]
  if (requireNamespace("matrixStats", quietly = TRUE)) {
    qs <- matrixStats::rowQuantiles(S, probs = c(0.025, p0, 0.975), na.rm = TRUE)
  } else {
    qs <- t(apply(S, 1L, quantile, probs = c(0.025, p0, 0.975), na.rm = TRUE))
  }
  x <- idx; yv <- y[idx]; lo <- qs[,1]; q <- qs[,2]; hi <- qs[,3]
  ylim <- range(c(yv, lo, hi), finite = TRUE)
  op <- par(mar = c(3.5, 4.2, 3, 1) + 0.1); on.exit(par(op), add = TRUE)
  plot(x, yv, type = "n", xlab = "index", ylab = "y",
       main = "Synthesized posterior predictive (last 40)", ylim = ylim)
  polygon(c(x, rev(x)), c(lo, rev(hi)), border = NA,
          col = adjustcolor("steelblue", alpha.f = 0.30))
  lines(x, q, lwd = 3, col = "orange3")
  points(x, yv, pch = 16, cex = 0.5)
  invisible(NULL)
}

# Ensures samples shape is T x ns
if (nrow(syn$draws) != length(y)) syn$draws <- t(syn$draws)

# Minimal zoom plot (base graphics)
pp_plot_zoom(y = y, samples = syn$draws, p0 = 0.5, last_n = 40L)

```

> **CRAN-safety.** All examples set a seed, use tiny data, finish in a few seconds, and keep the pure-R path by default. For speed and stability we fix **scale** and **skewness**.

## FAQ / Troubleshooting

* **It runs slowly.**
  Use short series (≤ 200), fix **scale**/**skewness**, and keep discount factors near but below one (≈ 0.96–0.99). Enable C++ bridges only if your toolchain supports them.

* **ELBO dips slightly—bug?**
  Small downward blips are expected from approximation/IS noise. Look for an overall upward trend; if not, simplify the model or adjust variance/discounts.

* **Synthesis seems unchanged.**
  Ensure your anchors are well-separated (e.g., 0.25–0.5–0.75), and verify that each anchor model converged (check ELBO tail). If anchors already respect monotonicity, isotonic adjustment may be minimal.

* **OpenMP not available.**
  That’s fine. It is optional. Everything runs serially; examples here use the pure-R path.

## How to cite

Barata, R., Prado, R., & Sansó, B. (2022). *Fast inference for time-varying quantiles via flexible dynamic models with application to the characterization of atmospheric rivers*. **Annals of Applied Statistics**, 16(1), 247–271. [https://doi.org/10.1214/21-AOAS1497](https://doi.org/10.1214/21-AOAS1497)

## License

MIT © The authors. See `LICENSE`.

## Getting help

Open an issue: [https://github.com/AntonioAPDL/exdqlm/issues](https://github.com/AntonioAPDL/exdqlm/issues)