---
title: "exdqlm — Extended Dynamic Quantile Linear Models"
output:
  rmarkdown::github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE, message = FALSE
)
```

<!-- badges: start -->

[![R-CMD-check](https://github.com/AntonioAPDL/exdqlm/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/AntonioAPDL/exdqlm/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

`exdqlm` fits **dynamic quantile** state-space models using the **extended Asymmetric Laplace** (exAL) error family. It targets time-series applications where the goal is to model **conditional quantiles** (not only means) while keeping a familiar state-space specification (design/evolution matrices and a state vector).
In **v0.4.0**, this CRAN update consolidates the internal 0.4/0.5/0.6 feature line: dynamic **LDVB** (`exdqlmLDVB()`), synthesis (`exdqlm_synthesize_from_draws()`), static regression/exAL routines (`regMod()`, `exal_static_LDVB()`, `exal_static_mcmc()`), reduced AL/DQLM modes (`dqlm.ind = TRUE`), and static shrinkage priors (`beta_prior = "ridge"` / `"rhs"` / `"rhs_ns"`), while keeping backend defaults conservative for CRAN stability.

> **Terminology (exAL)**
> We use **exAL** for the extended Asymmetric Laplace family throughout this package. It generalizes the standard AL by adding a skewness parameter, allowing for asymmetric tails. The standard AL is a special case with zero skewness. We refer to the generalized AL from Kotz et al. as **Kotz-GAL** to avoid confusion.

---

## Installation

CRAN (when available):

```r
install.packages("exdqlm")
```

Development (GitHub):

```r
# install.packages("pak")
pak::pak("AntonioAPDL/exdqlm")
```

## Quick start (≤ 10 lines)

Local-level model at a **single quantile** (the median). We fix **scale** and **skewness** to keep it fast and stable for CRAN; we keep the **pure-R** path.

```{r quick-start}
set.seed(1)
library(exdqlm)

T      <- 120
state  <- cumsum(rnorm(T, sd = 0.2))
y      <- state + rnorm(T, sd = 1.0)

model  <- list(FF = matrix(1), GG = matrix(1), m0 = 0, C0 = 100)
options(exdqlm.use_cpp_kf = FALSE, exdqlm.use_cpp_samplers = FALSE)

fit <- exdqlmISVB(
  y = y, p0 = 0.5, model = model, df = 0.98, dim.df = 1,
  fix.sigma = TRUE, sig.init = 1.0,
  fix.gamma = TRUE, gam.init = 0.0
)

tail(fit$diagnostics$elbo, 3)
```

## Core concepts (at a glance)

* **State-space skeleton**: *design* (`FF`) and *evolution* (`GG`) matrices with a prior for the **state vector** (`m0`, `C0`).
* **Quantile of interest**: `p0` (e.g., `0.1`, `0.5`, `0.9`).
* **exAL errors**: controlled by **scale** and **skewness**; fixing them often stabilizes small examples.
* **Discount factors**: `df` and `dim.df` control evolution per block (e.g., trend vs seasonality).
* **ELBO**: recorded at `fit$diagnostics$elbo` (weakly monotone up to importance-sampling noise).

## What’s new in v0.4.0

* **Dynamic LDVB algorithm** via `exdqlmLDVB()` for exDQLM fitting.
* **Synthesis helper** `exdqlm_synthesize_from_draws()` to combine posterior quantile-draw objects.
* **Static regression support** via `regMod()`, `exal_static_LDVB()`, and `exal_static_mcmc()`.
* **Reduced AL/DQLM paths** across dynamic and static APIs via `dqlm.ind = TRUE`.
* **Static shrinkage priors** in both static LDVB/MCMC via `beta_prior = "ridge"`, `"rhs"`, or `"rhs_ns"`.
* **LDVB transform helper** `transfn_exdqlmLDVB()` and expanded static object generics (`exal_mcmc`, `exal_ldvb`).
* **C++ backend controls** retained as optional: Kalman bridge default **TRUE**; builders, samplers,
  and post-predictive C++ paths default **FALSE**.
* **ELBO diagnostics** retained for iterative monitoring.

> For CI/CRAN-style runs, keep optional C++ builders/samplers/post-pred **FALSE** and set `exdqlm.use_cpp_kf = FALSE` for strict R-path reproducibility.

### Runtime options (summary)

| Option                    | Default | Effect                           | Use when…                                |
| ------------------------- | :-----: | -------------------------------- | ---------------------------------------- |
| `exdqlm.use_cpp_kf`       |  TRUE   | C++ Kalman filter bridge         | you have compilers/OpenMP and want speed |
| `exdqlm.use_cpp_builders` |  FALSE  | C++ matrix builders (`polytrendMod`, `seasMod`) | opt-in parity-tested builder speedups |
| `exdqlm.use_cpp_samplers` |  FALSE  | C++ samplers for posterior draws | same as above; keep OFF on CRAN/examples |
| `exdqlm.use_cpp_postpred` |  FALSE  | C++ posterior predictive sampler | optional speed path after parity checks  |
| `exdqlm.use_cpp_mcmc`     |  TRUE   | MCMC backend routing             | C++ FFBS by default for MCMC             |
| `exdqlm.cpp_mcmc_mode`    | `fast`  | MCMC mode (`strict`/`fast`)      | strict parity checks or fast C++ FFBS    |

Set with:

```r
options(exdqlm.use_cpp_kf = TRUE)
options(exdqlm.use_cpp_builders = FALSE)
options(exdqlm.use_cpp_samplers = TRUE)
options(exdqlm.use_cpp_postpred = FALSE)
options(exdqlm.use_cpp_mcmc = TRUE)
options(exdqlm.cpp_mcmc_mode = "fast")
```

Backend control (minimal):
* Force pure-R backend: set `options(exdqlm.use_cpp_kf = FALSE, exdqlm.use_cpp_builders = FALSE)`.
* Keep builder calls explicit with `backend = "R"` or `backend = "cpp"` in `polytrendMod()` and `seasMod()`.

## Minimal examples (CRAN-safe)

### 1) Single-quantile fit on built-in data (tiny slice)

Trend + seasonality + one regressor (`nino34`). **Note**: `FF` for the regressor is `1 × T`. Combine components **pairwise**.

```{r builtins-fit}
set.seed(2)
T <- 150
y <- log(BTflow[seq_len(T)])
x <- nino34[seq_len(T)]

trend.comp <- polytrendMod(order = 1, m0 = 0, C0 = 1)
seas.comp  <- seasMod(p = 12, h = 1, C0 = diag(1, 2))

# 1-d regressor block (explicit 1 x T design)
reg.comp <- list(m0 = 0, C0 = 1, FF = matrix(x, nrow = 1), GG = matrix(1))

# combine via +.exdqlm
reg.comp <- as.exdqlm(reg.comp)
model    <- trend.comp + seas.comp + reg.comp

# one discount per block: (trend, seasonal[2-d], reg)
df     <- c(1.00, 0.98, 1.00)
dim.df <- c(1,       2,   1)

options(exdqlm.use_cpp_kf = FALSE, exdqlm.use_cpp_samplers = FALSE)

fit <- exdqlmISVB(
  y = y, p0 = 0.5, model = model,
  df = df, dim.df = dim.df,
  fix.sigma = TRUE, sig.init = 0.2,
  fix.gamma = TRUE, gam.init = 0.0
)

# quick checks
tail(fit$diagnostics$elbo, 2)
dim(fit$theta.out$sm)  # state-dimension x time
```

### 2) exAL helper sanity check (CDF ↔ quantile)

```{r exal-helpers}
set.seed(3)
x      <- seq(-2, 2, length.out = 5)
p0     <- 0.25
mu     <- 0
sigma  <- 1
gamma  <- 0.0

# CDF then invert with QF — should approximately return x
cdf_vals <- pexal(x,  p0 = p0, mu = mu, sigma = sigma, gamma = gamma)
x_back   <- qexal(cdf_vals, p0 = p0, mu = mu, sigma = sigma, gamma = gamma)

round(cbind(x, x_back), 4)

# A few random draws
rexal(5, p0 = p0, mu = mu, sigma = sigma, gamma = gamma)
```

> **CRAN-safety.** All examples set a seed, use tiny data, finish in a few seconds, and explicitly keep the pure-R path.

### 3) Static reduced AL + RHS-family API sketch

```{r static-api-sketch, eval=FALSE}
set.seed(4)
n <- 80
p <- 5
X <- matrix(rnorm(n * p), n, p)
beta <- c(1, -1, 0, 0, 0.5)
y <- as.numeric(X %*% beta + rnorm(n))

# Reduced AL fit (gamma fixed at zero)
fit_al <- exal_static_LDVB(
  y = y, X = X, p0 = 0.5,
  dqlm.ind = TRUE,
  max_iter = 150, tol = 1e-4, verbose = FALSE
)

# exAL fit with regularized horseshoe prior on coefficients
fit_rhs <- exal_static_mcmc(
  y = y, X = X, p0 = 0.5,
  beta_prior = "rhs",
  n.burn = 200, n.mcmc = 200, thin = 1,
  mh.proposal = "slice",
  trace.diagnostics = FALSE,
  verbose = FALSE
)

# exAL fit with rhs_ns controls (same API family, additive option)
fit_rhs_ns <- exal_static_mcmc(
  y = y, X = X, p0 = 0.5,
  beta_prior = "rhs_ns",
  beta_prior_controls = list(
    tau0 = 0.5,
    a_zeta = 2,
    b_zeta = 1,
    shrink_intercept = FALSE
  ),
  n.burn = 200, n.mcmc = 200, thin = 1,
  mh.proposal = "slice",
  trace.diagnostics = FALSE,
  verbose = FALSE
)

fit_al$dqlm.ind
fit_rhs$beta_prior$type
fit_rhs_ns$beta_prior$type
```

## FAQ / Troubleshooting

* **It runs slowly.**
  Use short series (≤ 200), fix **scale**/**skewness**, and keep discount factors near but below one (≈ 0.96–0.99). Enable C++ bridges only if your toolchain supports them.

* **ELBO dips slightly - bug?**
  Small downward blips are expected from importance-sampling noise. Look for an overall upward trend. If not, simplify the model or adjust variance/discounts.

* **OpenMP not available.**
  That's fine. OpenMP is optional. The package compiles and runs serially; the examples here use the pure-R path.
* **Numerical stability tips.**
  Avoid extremely tight `C0`; start with moderate priors (e.g., `C0` around 1-100 for simple models), and fix **scale**/**skewness** for initial runs.

## How to cite

Barata, R., Prado, R., & Sansó, B. (2022). *Fast inference for time-varying quantiles via flexible dynamic models with application to the characterization of atmospheric rivers*. **Annals of Applied Statistics**, 16(1), 247–271. [https://doi.org/10.1214/21-AOAS1497](https://doi.org/10.1214/21-AOAS1497)

## License

MIT © The authors. See `LICENSE`.

## Getting help

Open an issue: [https://github.com/AntonioAPDL/exdqlm/issues](https://github.com/AntonioAPDL/exdqlm/issues)
