---
title: "exdqlm — Extended Dynamic Quantile Linear Models"
output:
  rmarkdown::github_document:
    html_preview: false
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE, message = FALSE
)
```

<!-- badges: start -->

[![R-CMD-check](https://github.com/AntonioAPDL/exdqlm/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/AntonioAPDL/exdqlm/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

`exdqlm` fits **dynamic quantile** state-space models using the **extended Asymmetric Laplace** (exAL) error family. It’s for time-series problems where you want conditional **quantiles**—not just means—while keeping a familiar state-space structure (design/evolution matrices and a state vector).
**v0.3.0** adds **optional C++ bridges** for speed (Kalman filter and samplers) and **ELBO monitoring** for the variational routine; defaults remain **pure-R** for portability.

> **Terminology (exAL)**
> We use **exAL** for the extended Asymmetric Laplace family throughout this package. It generalizes the standard AL by adding a skewness parameter, allowing for asymmetric tails. The standard AL is a special case with zero skewness. We refer to the generalized AL from Kotz et al. as **Kotz-GAL** to avoid confusion.

---

## Installation

CRAN (when available):

```r
install.packages("exdqlm")
```

Development (GitHub):

```r
# install.packages("pak")
pak::pak("AntonioAPDL/exdqlm")
```

---

## Quick start (≤ 10 lines)

A tiny example with a local-level model and a **single quantile** (the median). We fix **scale** and **skewness** to keep it fast and stable for CRAN.

```{r}
set.seed(1)
library(exdqlm)

T <- 120
state <- cumsum(rnorm(T, sd = 0.2))
y <- state + rnorm(T, sd = 1.0)

model <- list(FF = matrix(1), GG = matrix(1), m0 = 0, C0 = 100)

options(exdqlm.use_cpp_kf = FALSE, exdqlm.use_cpp_samplers = FALSE)

fit <- exdqlmISVB(
  y = y, p0 = 0.5, model = model, df = 0.98, dim.df = 1,
  fix.sigma = TRUE, sig.init = 1.0,
  fix.gamma = TRUE, gam.init = 0.0
)

tail(fit$diagnostics$elbo, 3)
```

---

## Core concepts (at a glance)

* **State-space skeleton**: specify *design* (`FF`) and *evolution* (`GG`) matrices and a prior for the **state vector** (`m0`, `C0`).
* **Quantile of interest**: `p0` (e.g., 0.1, 0.5, 0.9).
* **exAL errors**: controlled by **scale** and **skewness** (you can fix them for speed).
* **Discount factors**: `df` and `dim.df` control state evolution per block (e.g., trend vs seasonality).
* **ELBO**: tracked in `fit$diagnostics$elbo` for quick convergence checks (weakly monotone up to IS noise).

---

## What’s new in v0.3.0

* **C++ Kalman filter bridge** (drop-in):
  `options(exdqlm.use_cpp_kf = TRUE)` (default **FALSE**; examples keep it OFF).
* **ELBO monitoring** for IS-VB:
  per-iteration values in `fit$diagnostics$elbo`.
* **Optional C++ samplers** for posterior draws:
  `options(exdqlm.use_cpp_samplers = TRUE)` (default **FALSE**).
* **Portability**: OpenMP is **optional and gated**; package runs serially without it.

> Keep both options **FALSE** in examples/CI. Enable them locally only if your toolchain supports compiled code.

---

## Minimal examples (CRAN-safe)

### 1) Single-quantile fit on built-in data (tiny slice)

Trend + seasonality + one regressor (`nino34`). We fix **scale** and **skewness**.

```{r}
set.seed(2)
T <- 150
y <- log(BTflow[seq_len(T)])
x <- nino34[seq_len(T)]

trend.comp <- polytrendMod(order = 1, m0 = 0, C0 = 1)
seas.comp  <- seasMod(p = 12, h = 1, C0 = diag(1, 2))

# <-- define regressor as its own 1-d block
reg.comp <- list(m0 = 0, C0 = 1, FF = matrix(x, nrow = 1), GG = matrix(1))

# <-- combine pairwise
base.mod <- combineMods(trend.comp, seas.comp)
model    <- combineMods(base.mod, reg.comp)

# one discount per block: (trend, seasonal[2-d], reg)
df     <- c(1.00, 0.98, 1.00)
dim.df <- c(1,       2,   1)

options(exdqlm.use_cpp_kf = FALSE, exdqlm.use_cpp_samplers = FALSE)

fit <- exdqlmISVB(
  y = y, p0 = 0.5, model = model,
  df = df, dim.df = dim.df,
  fix.sigma = TRUE, sig.init = 0.2,
  fix.gamma = TRUE, gam.init = 0.0
)

# quick checks
tail(fit$diagnostics$elbo, 2)
dim(fit$theta.out$sm)  # state-dimension x time
```

### 2) exAL helper sanity check (CDF ↔ quantile)

```{r}
set.seed(3)
library(exdqlm)

x  <- seq(-2, 2, length.out = 5)
p0 <- 0.25
mu <- 0; sigma <- 1; gamma <- 0.0

# CDF then invert with QF — should approximately return x
cdf_vals <- pexal(x,  p0 = p0, mu = mu, sigma = sigma, gamma = gamma)
x_back   <- qexal(cdf_vals, p0 = p0, mu = mu, sigma = sigma, gamma = gamma)

round(cbind(x, x_back), 4)

# A few random draws
rexal(5, p0 = p0, mu = mu, sigma = sigma, gamma = gamma)
```

> **Tip (local only):**
> Try the C++ bridges by setting
>
> ```r
> options(exdqlm.use_cpp_kf = TRUE)
> options(exdqlm.use_cpp_samplers = TRUE)
> ```
>
> If your toolchain lacks OpenMP/compilers, the package automatically uses the pure-R path.

---

## FAQ / Troubleshooting

* **It runs slowly.**
  Use short series (≤ 200), fix **scale**/**skewness**, and keep discount factors near but below one (e.g., 0.96–0.99). Enable the C++ bridges only if your toolchain supports them.

* **OpenMP not available.**
  That's fine. OpenMP is optional. The package compiles and runs serially; the examples here use the pure-R path.

* **ELBO dips slightly - bug?**
  Small downward blips are expected from importance-sampling noise. Look for an overall upward trend. If not, simplify the model or adjust variance/discounts.

* **Numerical stability tips.**
  Avoid extremely tight `C0`; start with moderate priors (e.g., `C0` around 1-100 for simple models), and fix **scale**/**skewness** for initial runs.

---

## How to cite

Barata, R., Prado, R., & Sansó, B. (2022). *Fast inference for time-varying quantiles via flexible dynamic models with application to the characterization of atmospheric rivers*. **Annals of Applied Statistics**, 16(1), 247–271. [https://doi.org/10.1214/21-AOAS1497](https://doi.org/10.1214/21-AOAS1497)

---

## License

MIT © The authors. See `LICENSE`.

---

## Getting help

Please open an issue: [https://github.com/AntonioAPDL/exdqlm/issues](https://github.com/AntonioAPDL/exdqlm/issues)

---

**Reproducibility**: all examples set a seed, use tiny data, finish in a few seconds, and default to the pure-R path.

---
